---
title: "Do people who receive more years of education ultimately spend more hours working per week?"
author: "Aarna Sitani"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---


## Milestone #3

```{r}
library(ggplot2)
data <- read.csv("adult.data.csv")
names(data)[c(5,13)] <- c("education-num", "hours-per-week")
ggplot(data, aes(x = `education-num`, y = `hours-per-week`)) + 
  geom_point(alpha = 0.5) + 
  labs(title = "Relationship between Education and Hours Worked",
       x = "Years of Education",
       y = "Hours per Week") + 
  theme_minimal()

```


## Milestone #4

```{r}
ggplot(data, aes(x = as.factor(`education-num`), y = `hours-per-week`)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) + 
  labs(title = "Distribution of Hours Worked by Education Level",
       x = "Years of Education",
       y = "Hours per Week") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## FINAL REPORT:

## (1) Introduction: 

This study intends to answer the following question: do people who receive more years of education ultimately spend more hours working per week?  My data analysis aims to identify whether patterns exist between the duration individuals dedicate to schooling and the time they allocate to work post-education. My hypothesis posits a positive correlation between extended educational commitment and increased weekly working hours.  The underlying assumption is that individuals pursuing advanced academic paths, such as graduate studies, masters, and PhD programs, tend to be deeply committed to their work.  They are likely to invest additional time beyond the basic requirement of their field of study due to a genuine passion for their careers, viewing them not merely as means of income but as areas of profound interest.  Consequently, post-education, it seems plausible that these individuals continue to devote more time to their professional pursuits.  However, as almost hypocritical as it may sound given the previous argument, it is worth  acknowledging the potential confounding variable of income.  Many high-paying professions, such as those in medicine in law, necessitate ongoing education and, also, involve demanding working hours.  These lucrative careers could influence the observed positive correlation between educational duration and weekly working hours, as individuals in such fields may be required to invest significant time in their occupations for reasons beyond passion alone.  

Understanding and identifying the patterns in this study is interesting and important for several reasons.  Specifically, they could provide valuable insights for individuals planning their careers, helping them understand the potential time commitments associated with various levels of education.  This information can guide career decisions and expectations about work-life balance.  Employers and policymakers can benefit from insights into the relationship between education and working hours.  This knowledge can aid in workforce planning, identifying trends in labor market demands, and predicting future needs for skilled professionals.  


## (2) Data: 

This data set is sourced from the Machine Learning Repository of University of California, Irvine Census Income Data Set (UC Irvine).  The UCI page, though, mentions US Census Bureau as the original source of the data.  The independent variable in this report is the amount of years spent in school, and it is measured simply by count.  The dependent variable is the amount of hours worked weekly, and it is also measured by count. Both variables have been taken from this Census Bureau's data set.  

The Census Bureau uses data from a variety of sources.  Some are collected from respondents and businesses directly through censuses and surveys they conduct.  Others come from pre-existing agencies; such data, known as "administrative data" are reused to cut the cost of data collection and research, as well as reduce the burden on people who respond to their census and survey questions.  My research design is cross-sectional because the data are collected from participants at a specific point in time, and no manipulation or intervention is implied.  

```{r}
library(ggplot2)

ggplot(data, aes(x = `hours-per-week`)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Weekly Working Hours",
       x = "Hours per Week",
       y = "Frequency") +
  theme_minimal() +
    theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "white"),
        axis.line = element_line(color = "black"),
        axis.text = element_text(color = "steelblue"),
        axis.title = element_text(color = "steelblue", size = 14, face = "bold"),
        plot.title = element_text(hjust = 0.5, size = 18, face = "bold", color = "darkslategray"),
        plot.subtitle = element_text(hjust = 0.5, size = 14, color = "darkslategray"))
  
```
The plot visualizing the main outcome of interest here is fairly symmetric and does not particularly skew towards one end.  There is an extremely high frequency of people working around 40 hours a week, and the amount of people working at other hours tables off slowly as the data moves away from this center point.  


## (3) Results:

```{r}
## Boxplot
ggplot(data, aes(x = as.factor(`education-num`), y = `hours-per-week`)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) + 
  labs(title = "Distribution of Hours Worked by Education Level",
       x = "Years of Education",
       y = "Hours per Week",
       caption = "Source: US Census Bureau") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "darkgray"),
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position="none",
        plot.caption = element_text(hjust = 0, size = 10, color = "gray"))
```
Results (Interpreting Boxplots): 

At first glance, one may notice the relatively similar medians across different education levels centering around approximately 40 hours per week.  Where the greater differences come in, though, is the spread. In observing both the interquartile range and the whiskers of the plot, it becomes apparent that there is much variability in the range of weekly working hours.  The patterns of asymmetry and skewness is also an interesting factor to consider.  Because the distribution is relatively asymmetric, it seems that the mean and median are not necessarily of similar values; the mean may be influenced by extreme values.  Additionally, looking at the data up to about 8 years of education, the 75 percentile (Q3) mark is very close to the median value (Q2), but for the data above 8 years of education, the 25 percentile mark (Q1) is just touching the median value.  


```{r}
## Violin Plot
library(ggplot2)

ggplot(data, aes(x = as.factor(`education-num`), y = `hours-per-week`)) + 
  geom_violin(fill = "skyblue", alpha = 0.7) + 
  labs(title = "Distribution of Hours Worked by Education Level", 
       x = "Years of Education", 
       y = "Hours per Week", 
       caption = "Source: US Census Bureau") +
  theme_minimal()
```
Results (Interpreting Violin Plot):

While we did not cover violin plots in class, I thought it would be interesting to provide deeper insight into the distribution of hours worked for each level of education, as it combines aspects of box plots and kernel density plots to visualize the density of the data. Comparing the location of the figures here against the data from the box plot leads to unsurprising findings, for this visualization too reveals a high frequency of 40 hours worked per week.  The distribution of hours per week at 9 years of education, though, is particularly fascinating; there is much more variation at this amount of education compared to the other amounts of time spent in school. 


```{r}
## Regression
library(ggplot2)

model <- lm(`hours-per-week` ~ `education-num`, data = data)

ggplot(data, aes(x = `education-num`, y = `hours-per-week`)) + 
  geom_point(alpha = 0.7, color = "steelblue", size = 1.5) + 
  geom_smooth(method = "lm", se = FALSE, color = "red", size = 1, linetype = "dashed") + 
  labs(title = "Relationship between Education and Hours Worked", 
       x = "Years of Education",
       y = "Hours per Week",
       caption = "Source: US Census Bureau") + 
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold", color = "darkblue"),
        axis.title = element_text(size = 14, color = "darkblue"),
        axis.text = element_text(size = 12, color = "darkblue"),
        plot.caption = element_text(hjust = 0, size = 10, color = "gray"))
```
```{r}
library(modelsummary)

model <- lm(`hours-per-week` ~ `education-num`, data = data)

modelsummary(model, stars = TRUE, tidy_names = list(`education-num` = "Years of Education"))
```


```{r}
library(kableExtra)

model <- lm(`hours-per-week` ~ `education-num`, data = data)
summary_table <- as.data.frame(summary(model)$coefficients)
kable(summary_table, "html") |>
  kable_styling()
```

Results (Interpreting Regression): 

The coefficient for `education-num` is 0.71093.  This represents the estimated change in weekly working hours for each additional year of education.  In other words, for every extra year of schooling, individuals are estimated to work approximately 0.71 more hours per week.  The p-value associated with the coefficient is very close to zero (p-value: < 2e-16), indicating that the relationship between education duration and weekly working hours is statistically significant.  Said otherwise, the likelihood of observing such a relationship due to random change is extremely low.  The residuals provide information about the differences between the observed and predicted values.  The spread of residuals (ranging from -43.646 to 62.885) indicates the variability in weekly working hours that is not explained by the model.  The Multiple R-squared is 0.02194, which suggests that the model explains approximately 2.19% of the variance in weekly working hours.  This means that while the model is statistically significant, it does not account for a large proportion of variability in the data.  Determining causation in general from a regression model alone can be challenging.  While the coefficient here indicates an association, declaring causation requires additional evidence, such as experimental design or controlling for potential confounding variables.  Therefore, it would be inappropriate to conclude a causal effect.  

## (4) Conclusion:
In this study, I explored the relationship between the duration of education and the number of hours individuals dedicate to work per week.  The regression analysis supported my hypothesis, indicating a statistically significant positive association between years of education and weekly working hours.  The coefficient of 0.71093 suggested that, on average, individuals can be expected to work approximately 0.71 more hours per week for each additional year of education.  While statistically significant, the model explained only a modest proportion (2.19%) of the variance in weekly working hours.  Despite the insights gained, several limitations and threats to inference should be acknowledged.  Firstly, the cross-sectional nature of the study restricts our ability to establish causation.  The observed association does not necessarily imply a causal relationship, and other factors may contribute to the patterns identified.  The potential confounding variable of income poses a significant challenge to the analysis.  High-paying professions, which often require extended education, may influence the positive correlation observed.  Future research could benefit from controlling for income levels to better isolate the impact of education on working hours.  Additionally, the limited explanatory power of the model (2.19% variance explained) suggests that other unexplored factors contribute to the variability in working hours.  These factors might include individual preferences, industry-specific demands, or external life circumstances.  Missing data or measurement errors in the data set could also introduce bias to the analysis.  Given more time and resources, several enhancements to the analysis could be pursued.  Firstly, a longitudinal study design would allow for a more nuanced exploration of the dynamic relationship between education and working hours over time.  Controlling for additional variables, such as income and job industry, would strengthen the analysis by addressing potential confounding variables.  Moreover, qualitative research methods, such as interviews of surveys, could provide deeper insights into individuals' motivations and perceptions regarding their work commitments post-education. 