---
title: "Gov 50 Final Project"
author: "Aarna Sitani"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to...
analyzing political campaign finance, customer behavior in online retail, patterns of employee productivity, social media influence on brand awareness


## Milestone #2: What is the relationship between the years of educational attainment and the number of hours worked per week?  My data analysis intends to identify whether there are any patterns between the time people spend in school and the time they spend working post-education.  My hypothesis is that those who spend more time in school end up spending more hours working per week — essentially a positive correlation between the two variables.  This is because I would assume that those who continue their academic career beyond college into graduate schools, masters, and PhD programs find themselves deeply attached to their work and are willing to pursue time additional to what is necessary in their field of study.  This willingness to spend extra time likely is driven from seeing their careers as an area of genuine passion, not just a way to make money, so it would make sense that post-education, these people would continue investing more of their time in working.  That being said, though, there is also the potential confounding variable of income.  Oftentimes, high-paying jobs require continued education, like medicine school or law, and these professions also have demanding working hours.  For the study, my explanatory variable is the number of years of education, and it will be coded as =[number of years of education].  More specifically, if the variable were coded as =9, that would mean, at one data point, someone spent 9 years in school.  My outcome variable is the number of hours worked per week, and it will be coded as =[number of hours worked per week].  If after running the regression analysis, the data reveals a positive, significant coefficient, I would receive evidence supporting my hypothesis.  If there results a negative coefficient from my data, the evidence would disprove my hypothesis. 




## Milestone #3

```{r}
library(ggplot2)
data <- read.csv("adult.data.csv")
names(data)[c(5,13)] <- c("education-num", "hours-per-week")
ggplot(data, aes(x = `education-num`, y = `hours-per-week`)) + 
  geom_point(alpha = 0.5) + 
  labs(title = "Relationship between Education and Hours Worked",
       x = "Years of Education",
       y = "Hours per Week") + 
  theme_minimal()

```


## Milestone #4

```{r}
ggplot(data, aes(x = as.factor(`education-num`), y = `hours-per-week`)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) + 
  labs(title = "Distribution of Hours Worked by Education Level",
       x = "Years of Education",
       y = "Hours per Week") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## FINAL REPORT:

## (1) Introduction: This study intends to answer the following question: do people who receive more years of education ultimately spend more hours working per week?  My data analysis aims to identify whether patterns exist between the duration individuals dedicate to schooling and the time they allocate to work post-education. My hypothesis posts a positive correlation between extended educational commitment and increased weekly working hours.  The underlying assumption is that individuals pursuing advanced academic paths, such as graduate studies, masters, and PhD programs, tend to be deeply committed to their work.  They are likely to invest additional time beyond the basic requirement of their field of study due to a genuine passion for their careers, viewing them not merely as means of income but as areas of profound interest.  Consequently, post-education, it seems plausible that these individuals continue to devote more time to their professional pursuits.  However, as almost hypocritical as it may sound given the previous argument, it is worth  acknowledging the potential confounding variable of income.  Many high-paying professions, such as those in medicine in law, necessitate ongoing education and, also, involve demanding working hours.  These lucrative careers could influence the observed positive correlation between educational duration and weekly working hours, as individuals in such fields may be requied to invest significant time in their occuptions for reasons beyond passion alone.  Understanding and identifying the patterns in this study is interesting and important for several reasons.  Specifically, they could provide valuable insights for individuals planning their careers, helping them understand the potential time commitments associated with various levels of education.  This information can guide career decisions and expectations about work-life balance.  Employers and policymakers can benefit from insights into the relationship between education and working hours.  This knowledge can aid in workforce planning, identifying trends in labor market demands, and predicting future needs for skilled professionals.  


## (2) Data: This data set is sourced from the Machine Learning Repository of University of California, Irvine Census Income Data Set (UC Irvine).  The UCI page, though, mentions US Census Bureau as the original source of the data.  The Census Bureau uses data from a variety of sources.  Some are collected from respondents and businesses directly through censuses and surveys they conduct.  Others come from pre-existing agencies; such data, known as "administrative data" are reused to cut the cost of data collection and research, as well as reduce the burden on people who respond to their census and survey questions.  My research design is cross-sectional because the data are collected from participants at a specific point in time, and no manipulation or intervention is implied.  

```{r}
ggplot(data, aes(x = `hours-per-week`)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Weekly Working Hours",
       x = "Hours per Week",
       y = "Frequency") +
  theme_minimal()
```


## (3) Results:

```{r}
## Boxplot
ggplot(data, aes(x = as.factor(`education-num`), y = `hours-per-week`)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) + 
  labs(title = "Distribution of Hours Worked by Education Level",
       x = "Years of Education",
       y = "Hours per Week") + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
## Results (Interpreting Boxplots): At first glance, one may notice the relatively similar medians across different education levels centering around approximately 40 hours per week.  Where the greater differences come in, though, is the spread. In observing both the interquartile range and the whiskers of the plot, it becomes apparent that there is much variability in the range of weekly working hours.  The patterns of asymmetry and skewness is also an interesting factor to consider.  Because the distribution is relatively asymmetric, it seems that the mean and median are not necessarily of similar values; the mean may be influenced by extreme values.  Additionally, looking at the data up to about 8 years of education, the 75 percentile (Q3) mark is very close to the median value (Q2), but for the data above 8 years of education, the 25 percentile mark (Q1) is just touching the median value.  

```{r}
## Regression
library(ggplot2)

model <- lm(`hours-per-week` ~ `education-num`, data = data)

ggplot(data, aes(x = `education-num`, y = `hours-per-week`)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE, color = "red", formula = y ~ x) + 
  labs(title = "Relationship between Education and Hours Worked", 
       x = "Years of Education",
       y = "Hours per Week") + 
  theme_minimal()
```
## Results (Interpreting Regression): The coefficient for `education-num` is 0.71093.  This represents the estimated change in weekly working hours for each additional year of education.  In other words, for every extra year of schooling, individuals are etimated to work approximately 0.71 more hours per week.  The p-value associated with the coefficient is very close to zero (p-value: < 2e-16), indicating that the relationship between education duration and weekly working hours is statistically significant.  Said otherwise, the likelihood of observing such a relationship due to random change is extremely low.  The residuals provide information about the differences between the observed and predicted values.  The spread of residuals (ranging from -43.646 to 62.885) indicates the variability in weekly working hours that is not explained by the model.  The Multiple R-squared is 0.02194, qhich suggests that the model explains approximately 2.19% of the variance in weekly working hours.  This means that while the model is statistically significant, it does not account for a large proportion of variability in the data.  Determining causation in general from a regression model alone can be challenging.  While the coefficient here indicates an association, declaring causation requires addiional evidence, such as experimental design or controlling for potential confounding variables.  Therefore, it would be inappropriate to conclude a causal effect.  

## (4) Conclusion: